---
title: "Infection in heterogeneous population with interventions"
author: "C. Bartels"
date: "18 April 2020"
output: word_document
---

## simulations

```{r}
# setup
require(dplyr)
require(tidyr)
require(ggplot2)
require(gganimate)
set.seed(123)

theme_set(theme_classic())

# size of population
nPop <- 10^5


# end of simulation
tend <- 365

# infective period, assumed to start one day after infection with duration
idur <- 7

# number of repeated simulations
nsim <- 50

# homgenous population
  # rate of new infections per day 
  r0a = 0.4
  # between subject variability of r0
  sd0a = 0.0

# heterogenous population
  # rate of new infections per day
  r0b = 0.2
  # between subject variability of r0
  sd0b = 1.0



# simulate spread of infection
mySim <- function(pat1, int1 = NULL, rVac = 0.0) {
  
  # number of patients that get vaccinated
  nVac <- max(1.0, min(nrow(pat1), rVac*nrow(pat1)))
  
  pat1 <-
    pat1 %>%
    mutate(
      is = 0, # day of infection (0 ... not infected)
      ip = 1:n(), # subject identifier
      is = ifelse(ip <= nVac, 99999, is), # vacinated subjects
      is = ifelse(ip %in% ceiling(runif(10,nVac,n())), 1, is), # 10 random subjects get infected
      cru = cumsum(r0) / sum(r0) # to select subjets that get infected with probability proportional to r0
    )

  # simulate day by day
  for(iw in 2:tend) { # iw <- 2
    
    # intervention
    if(is.null(int1)) {
      if1 <- 1
    } else {
      if1 <- int1 %>% filter(t == iw) %>% select(rf) %>% unlist()
    }
  
    # select patients that are infective
    pat2 <- pat1 %>% filter(is != 0 & is > iw-idur & is <= iw)
  
    # rate of transmission
    rt <- sum(pat2$r0) * if1
    
    # number of subjects infected
    ni <- rpois(1, rt)
    
    if(ni > 0) {
      # randomly select the patients who get transmitted
      pt1 <- runif(ni, min = 0.0, max = 1.0)
      ip1 <- findInterval(pt1, pat1$cru) + 1
      
      # select patients who receive infection
      pat1$sel <- FALSE
      pat1[ip1, "sel"] <- TRUE
  
      # infect patients who are not immune
      pat1 <- 
        pat1 %>%
        mutate(is = ifelse(is == 0 & sel, iw, is))
    }
    
  }

  return(pat1)
}

# repeated simulations and summarize as median and percentile
mySim2 <- function(r0, sd0, int1 = NULL) 
{
  
  # subjects
  pat1 <- tibble(
    r0l = rnorm(nPop, log(r0), sd0), # log of patient specific rate
    r0 = exp(r0l) / exp(sd0^2/2) # patient specific rate (rescaled to have mean r0)
  )

  sims <- list()
  for(isim in 1:nsim) {
    
    sim1 <- mySim(pat1, int1)

    pp1 <-
      ggplot(sim1, aes(x=r0, fill=(is!=0))) + 
      geom_density(alpha=0.4, position="stack") + 
      scale_x_log10()
  
    s2 <- sim1 %>% group_by(is) %>% summarise(ni = n())
  
    ts1 <- 
      tibble(t = 1:tend, isim=isim) %>% 
      left_join(s2, by = c("t" = "is")) %>% 
      mutate(
        ni = ifelse(is.na(ni), 0, ni),
        nt = cumsum(ni)) 
    
    sims[[isim]] <- ts1
  
  }
  
  sim2 <- 
    bind_rows(sims) %>%
    group_by(t) %>%
    summarise(
      ni25 = quantile(ni, probs = 0.25),
      nim = median(ni),
      ni75 = quantile(ni, probs = 0.75),
      nt25 = quantile(nt, probs = 0.25),
      ntm = median(nt),
      nt75 = quantile(nt, probs = 0.75)
    )

}


# Illustrate heterogenous patient population
pat1 <- tibble(
  r0l = rnorm(nPop, log(r0b), sd0b), # log of patient specific rate
  r0 = exp(r0l) / exp(sd0b^2/2) # patient specific rate (rescaled to have mean r0)
)

p0 <-
  ggplot(pat1, aes(x=r0)) + 
  geom_density() +
  scale_x_log10() +
  xlab("Infection rate (1/Day)")

print(p0)

# make animated graph of disease progression
mygraph1 <- function(sim1) {
  sim1 <-
    sim1 %>%
    mutate(bin = cut_interval(log(r0), n = 40)) %>%
    group_by(bin) %>%
    mutate(
      dbin = median(r0),
      order = as.double(bin)) %>%
    ungroup()
  
  an1 <-
    sim1 %>%
    expand(ip, iweek) %>%
    rename(lweek = iweek) %>%
    full_join(sim1, by="ip") %>%
    ungroup()
  
  an2 <-
    an1 %>%
    group_by(dbin, bin, order, lweek) %>%
    summarise(
      n = sum(is>0 & iweek<=lweek),
      ntot = n(),
      ninf = sum(is>0),
      reff = mean(r0 * (is == 0 | iweek>lweek)) * (idur - 1)) %>%
    ungroup() %>%
    arrange(lweek, dbin)
  
  mylabs1 <- 
    sim1 %>% 
    select(order, dbin) %>% 
    unique() %>%
    mutate(lab = ifelse((order+2) %% 5 == 0, paste(signif(dbin,2)), ""))
  
  # Animation
  p <-
    an2 %>% 
    ggplot(aes(x = order)) +
    geom_bar(aes(y=ntot), stat = "identity", fill = "grey20", alpha=0.2) +
    geom_bar(aes(y=ninf), stat = "identity", fill = "red", alpha=0.2) +
    geom_bar(aes(y=n), stat = "identity", fill = "#ff9933") +
    labs(title=paste0('Week {closest_state}\nR(1 year) = ', round(R360, 2)), 
         x="Infection rate (1/Day)",
         y="Number of subjects") +
    theme(plot.title = element_text(hjust = 0.5, size = 18)) +
    scale_x_continuous(breaks=mylabs1$order, labels=mylabs1$lab) +
    transition_states(lweek, transition_length = 1, state_length = 5) +
    view_follow(fixed_y=TRUE, fixed_x = TRUE) +
    ease_aes('cubic-in-out')

  return(p)
}


```

## No intervention, no Heterogeneity

```{r}



sim2 <- mySim2(r0a, sd0a)
  
p1 <-
  ggplot(sim2, aes(x = t, y = ntm / nPop * 100)) +
  geom_ribbon(aes(ymin = nt25 / nPop * 100, ymax = nt75 / nPop * 100), alpha = 0.3) +
  geom_line() +
  xlab("Day") +
  ylab("% infected")

print(p1 + coord_cartesian(ylim = c(0, 100)))
print(p1 + scale_y_log10() + coord_cartesian(ylim = c(1, 100)))
print(p1 + scale_y_log10() + coord_cartesian(xlim = c(10,30)))

p2 <-
  ggplot(sim2, aes(x = t, y = nim / nPop * 100)) +
  geom_ribbon(aes(ymin = ni25 / nPop * 100, ymax = ni75 / nPop * 100), alpha = 0.3) +
  geom_line() +
  xlab("Day") +
  ylab("% infectious")

print(p2)


ts2 <- sim2 %>% mutate(Scenario = "Base")

```

## No intervention, with Heterogeneity

```{r}


sim2 <- mySim2(r0b, sd0b)
  

p1 <-
  ggplot(sim2, aes(x = t, y = ntm / nPop * 100)) +
  geom_ribbon(aes(ymin = nt25 / nPop * 100, ymax = nt75 / nPop * 100), alpha = 0.3) +
  geom_line() +
  xlab("Day") +
  ylab("% infected")

print(p1 + coord_cartesian(ylim = c(0, 100)))

p2 <-
  ggplot(sim2, aes(x = t, y = nim / nPop * 100)) +
  geom_ribbon(aes(ymin = ni25 / nPop * 100, ymax = ni75 / nPop * 100), alpha = 0.3) +
  geom_line() +
  xlab("Day") +
  ylab("% infectious")

print(p2)


ts2 <- bind_rows(
  ts2,
  sim2 %>% mutate(Scenario = "Heterogeneity")
)  

p1 <-
  ggplot(ts2 %>% filter(Scenario %in% c("Base", "Heterogeneity")), aes(x = t, y = ntm / nPop * 100)) +
  geom_line(aes(linetype = Scenario)) +
  geom_ribbon(data = ts2 %>% filter(Scenario == "Base"),
                aes(ymin = nt25 / nPop * 100, ymax = nt75 / nPop * 100), alpha = 0.3) +
  xlab("Day") +
  ylab("% infected")

print(p1 + coord_cartesian(ylim = c(0, 100)))
print(p1 + coord_cartesian(xlim = c(4, 30)) + scale_y_log10())


```



## Intervention from day 20 to 50, no Heterogeneity

```{r}


# intervention
int1 <- 
  tibble(t = 1:tend) %>%
  mutate(rf = ifelse(t>20 & t<50, 0.3, 1.0)) # period of social distancing


sim2 <- mySim2(r0a, sd0a, int1)
  

p1 <-
  ggplot(sim2, aes(x = t, y = ntm / nPop * 100)) +
  geom_ribbon(aes(ymin = nt25 / nPop * 100, ymax = nt75 / nPop * 100), alpha = 0.3) +
  geom_line() +
  xlab("Day") +
  ylab("% infected")

print(p1 + coord_cartesian(ylim = c(0, 100)))

p2 <-
  ggplot(sim2, aes(x = t, y = nim / nPop * 100)) +
  geom_ribbon(aes(ymin = ni25 / nPop * 100, ymax = ni75 / nPop * 100), alpha = 0.3) +
  geom_line() +
  xlab("Day") +
  ylab("% infectious")

print(p2)


ts2 <- bind_rows(
  ts2,
  sim2 %>% mutate(Scenario = "Intervention")
)  

p1 <-
  ggplot(ts2 %>% filter(Scenario %in% c("Base", "Intervention")), aes(x = t, y = ntm / nPop * 100)) +
  geom_line(aes(linetype = Scenario)) +
  geom_ribbon(data = ts2 %>% filter(Scenario == "Base"),
                aes(ymin = nt25 / nPop * 100, ymax = nt75 / nPop * 100), alpha = 0.3) +
  xlab("Day") +
  ylab("% infected")

print(p1 + coord_cartesian(ylim = c(0, 100)))


```


## Intervention from day 20 to 50, with Heterogeneity

```{r}


sim2 <- mySim2(r0b, sd0b, int1)
  

p1 <-
  ggplot(sim2, aes(x = t, y = ntm / nPop * 100)) +
  geom_ribbon(aes(ymin = nt25 / nPop * 100, ymax = nt75 / nPop * 100), alpha = 0.3) +
  geom_line() +
  xlab("Day") +
  ylab("% infected")

print(p1 + coord_cartesian(ylim = c(0, 100)))

p2 <-
  ggplot(sim2, aes(x = t, y = nim / nPop * 100)) +
  geom_ribbon(aes(ymin = ni25 / nPop * 100, ymax = ni75 / nPop * 100), alpha = 0.3) +
  geom_line() +
  xlab("Day") +
  ylab("% infectious")

print(p2)

ts2 <- bind_rows(
  ts2,
  sim2 %>% mutate(Scenario = "Both")
)  

p1 <-
  ggplot(ts2 %>% filter(Scenario %in% c("Heterogeneity", "Both")), aes(x = t, y = ntm / nPop * 100)) +
  geom_line(aes(linetype = Scenario)) +
  geom_ribbon(data = ts2 %>% filter(Scenario == "Both"),
                aes(ymin = nt25 / nPop * 100, ymax = nt75 / nPop * 100), alpha = 0.3) +
  xlab("Day") +
  ylab("% infected")

print(p1 + coord_cartesian(ylim = c(0, 100)))



```

## Intervention from day 20 to 50 with follow up to day 80, with Heterogeneity

```{r}


# intervention
int2 <- 
  tibble(t = 1:tend) %>%
  mutate(rf = case_when(
    t <= 20 ~ 1,
    t <  50 ~ 0.3,
    t <  80 ~ 0.6,
    TRUE ~ 1
  )) # period of social distancing

sim2 <- mySim2(r0b, sd0b, int2)
  
p1 <-
  ggplot(sim2, aes(x = t, y = ntm / nPop * 100)) +
  geom_ribbon(aes(ymin = nt25 / nPop * 100, ymax = nt75 / nPop * 100), alpha = 0.3) +
  geom_line() +
  xlab("Day") +
  ylab("% infected")

print(p1 + coord_cartesian(ylim = c(0, 100)))

p2 <-
  ggplot(sim2, aes(x = t, y = nim / nPop * 100)) +
  geom_ribbon(aes(ymin = ni25 / nPop * 100, ymax = ni75 / nPop * 100), alpha = 0.3) +
  geom_line() +
  xlab("Day") +
  ylab("% infectious")

print(p2)

ts2 <- bind_rows(
  ts2,
  sim2 %>% mutate(Scenario = "Follow up to day 80")
)  

p1 <-
  ggplot(ts2 %>% filter(Scenario %in% c("Follow up to day 80", "Both")), aes(x = t, y = ntm / nPop * 100)) +
  geom_line(aes(linetype = Scenario)) +
  geom_ribbon(data = ts2 %>% filter(Scenario == "Both"),
                aes(ymin = nt25 / nPop * 100, ymax = nt75 / nPop * 100), alpha = 0.3) +
  xlab("Day") +
  ylab("% infected")

print(p1 + coord_cartesian(ylim = c(0, 100)))



```

## Intervention from day 20 to 50 with follow up to day 180, with Heterogeneity

```{r}

# intervention
int3 <- 
  tibble(t = 1:tend) %>%
  mutate(rf = case_when(
    t <= 20 ~ 1,
    t <  50 ~ 0.3,
    t < 180 ~ 0.6,
    TRUE ~ 1
  )) # period of social distancing

sim2 <- mySim2(r0b, sd0b, int3)
  
p1 <-
  ggplot(sim2, aes(x = t, y = ntm / nPop * 100)) +
  geom_ribbon(aes(ymin = nt25 / nPop * 100, ymax = nt75 / nPop * 100), alpha = 0.3) +
  geom_line() +
  xlab("Day") +
  ylab("% infected")

print(p1 + coord_cartesian(ylim = c(0, 100)))

p2 <-
  ggplot(sim2, aes(x = t, y = nim / nPop * 100)) +
  geom_ribbon(aes(ymin = ni25 / nPop * 100, ymax = ni75 / nPop * 100), alpha = 0.3) +
  geom_line() +
  xlab("Day") +
  ylab("% infectious")

print(p2)

ts2 <- bind_rows(
  ts2,
  sim2 %>% mutate(Scenario = "Follow up to day 180")
)  

  
p1 <-
  ggplot(ts2 %>% filter(Scenario %in% c("Follow up to day 180", "Both")), aes(x = t, y = ntm / nPop * 100)) +
  geom_line(aes(linetype = Scenario)) +
  geom_ribbon(data = ts2 %>% filter(Scenario == "Follow up to day 180"),
                aes(ymin = nt25 / nPop * 100, ymax = nt75 / nPop * 100), alpha = 0.3) +
  xlab("Day") +
  ylab("% infected")

print(p1 + coord_cartesian(ylim = c(0, 100)))


```


## Intervention from day 20 to 50 with pronounced follow up to day 180, with Heterogeneity

```{r}


# intervention
int4 <- 
  tibble(t = 1:tend) %>%
  mutate(rf = case_when(
    t <= 20 ~ 1,
    t <  50 ~ 0.3,
    t < 180 ~ 0.5,
    TRUE ~ 1
  )) # period of social distancing

sim2 <- mySim2(r0b, sd0b, int4)
  
p1 <-
  ggplot(sim2, aes(x = t, y = ntm / nPop * 100)) +
  geom_ribbon(aes(ymin = nt25 / nPop * 100, ymax = nt75 / nPop * 100), alpha = 0.3) +
  geom_line() +
  xlab("Day") +
  ylab("% infected")

print(p1 + coord_cartesian(ylim = c(0, 100)))

p2 <-
  ggplot(sim2, aes(x = t, y = nim / nPop * 100)) +
  geom_ribbon(aes(ymin = ni25 / nPop * 100, ymax = ni75 / nPop * 100), alpha = 0.3) +
  geom_line() +
  xlab("Day") +
  ylab("% infectious")

print(p2)

ts2 <- bind_rows(
  ts2,
  sim2 %>% mutate(Scenario = "Pronounced follow up to day 180")
)  

  


```

## Intervention from day 20 to day 180, with Heterogeneity

```{r}


# intervention
int5 <- 
  tibble(t = 1:tend) %>%
  mutate(rf = case_when(
    t <= 20 ~ 1,
    t <  50 ~ 0.3,
    t < 180 ~ 0.3,
    TRUE ~ 1
  )) # period of social distancing

sim2 <- mySim2(r0b, sd0b, int5)
  
p1 <-
  ggplot(sim2, aes(x = t, y = ntm / nPop * 100)) +
  geom_ribbon(aes(ymin = nt25 / nPop * 100, ymax = nt75 / nPop * 100), alpha = 0.3) +
  geom_line() +
  xlab("Day") +
  ylab("% infected")

print(p1 + coord_cartesian(ylim = c(0, 100)))

p2 <-
  ggplot(sim2, aes(x = t, y = nim / nPop * 100)) +
  geom_ribbon(aes(ymin = ni25 / nPop * 100, ymax = ni75 / nPop * 100), alpha = 0.3) +
  geom_line() +
  xlab("Day") +
  ylab("% infectious")

print(p2)

ts2 <- bind_rows(
  ts2,
  sim2 %>% mutate(Scenario = "Intervention to day 180")
)  

  
p1 <-
  ggplot(ts2, aes(x = t, y = ntm / nPop * 100)) +
  geom_line(aes(colour = Scenario)) +
  geom_ribbon(data = ts2 %>% filter(Scenario == "Follow up to day 180"),
                aes(ymin = nt25 / nPop * 100, ymax = nt75 / nPop * 100), alpha = 0.3) +
  xlab("Day") +
  ylab("% infected")

print(p1 + coord_cartesian(ylim = c(0, 100)))
print(p1 + scale_y_log10() + coord_cartesian(ylim = c(1, 100)))
print(p1 + scale_y_log10() + coord_cartesian(xlim = c(4,30)))

p1 <-
  ggplot(ts2, aes(x = t, y = nim / nPop * 100, colour = Scenario)) +
  geom_line(aes(colour = Scenario)) +
  geom_ribbon(data = ts2 %>% filter(Scenario == "Follow up to day 180"),
                aes(ymin = ni25 / nPop * 100, ymax = ni75 / nPop * 100), alpha = 0.3) +
  xlab("Day") +
  ylab("% infectious")

print(p1 )

```

# plots of density
```{r}

# subjects
set.seed(123)
pat0 <- tibble(
  r0l = rnorm(nPop, log(r0a), 0.001), # log of patient specific rate
  r0 = exp(r0l) / exp(0.001^2/2) # patient specific rate (rescaled to have mean r0)
)

set.seed(123)
sim1 <- mySim(pat0)

R360 = sim1 %>% summarise(mr = mean(r0 * (is == 0)) * (idur - 1)) %>% unlist()


sim1 <-
  sim1 %>%
  mutate(
    iweek = pmin((is + 6) %/% 7, 52))

pp1 <-
  ggplot(sim1, aes(x=r0)) + 
  # geom_freqpoly(fill = NA, size=2) +
  geom_histogram(aes(fill = (is>0)), bins=40, alpha=0.2) +
  geom_freqpoly(aes(colour=factor(iweek)), position="stack", bins=40) + 
  scale_x_log10() +
  xlab("Infection rate (1/Day)") +
  scale_fill_discrete(labels = c("No", "Yes")) +
  guides(colour = "none", fill = guide_legend("Infected")) +
  ggtitle(paste0("Homogeneous\nR(1 year) = ", round(R360, 2)))


print(pp1)

p <- mygraph1(sim1)
animate(p, nframes=50, fps=4)
anim_save("animHom.gif", p)


# including vaccination
set.seed(123)
sim1 <- mySim(pat0, rVac = 1-1/r0a/(idur - 1))

R360 = sim1 %>% summarise(mr = mean(r0 * (is == 0)) * (idur - 1)) %>% unlist()


sim1 <-
  sim1 %>%
  mutate(
    iweek = pmin((is + 6) %/% 7, 52))

pp1 <-
  ggplot(sim1, aes(x=r0)) + 
  # geom_freqpoly(fill = NA, size=2) +
  geom_histogram(aes(fill = (is>0)), bins=40, alpha=0.2) +
  geom_freqpoly(aes(colour=factor(iweek)), position="stack", bins=40) + 
  scale_x_log10() +
  xlab("Infection rate (1/Day)") +
  scale_fill_discrete(labels = c("No", "Yes")) +
  guides(colour = "none", fill = guide_legend("Infected")) +
  ggtitle(paste0("Vaccination\nR(1 year) = ", round(R360, 2)))


print(pp1)

p <- mygraph1(sim1)
animate(p, nframes=50, fps=4)
anim_save("animHomVac.gif", p)


# intervention
set.seed(123)
sim1 <- mySim(pat0, int4)

R360 = sim1 %>% summarise(mr = mean(r0 * (is == 0)) * (idur - 1)) %>% unlist()


sim1 <-
  sim1 %>%
  mutate(
    iweek = pmin((is + 6) %/% 7, 52))

pp1 <-
  ggplot(sim1, aes(x=r0)) + 
  # geom_freqpoly(fill = NA, size=2) +
  geom_histogram(aes(fill = (is>0)), bins=40, alpha=0.2) +
  geom_freqpoly(aes(colour=factor(iweek)), position="stack", bins=40) + 
  scale_x_log10() +
  xlab("Infection rate (1/Day)") +
  scale_fill_discrete(labels = c("No", "Yes")) +
  guides(colour = "none", fill = guide_legend("Infected")) +
  ggtitle(paste0("Homogeneous with follow up to day 180\nR(1 year) = ", round(R360, 2)))


print(pp1)

p <- mygraph1(sim1)
animate(p, nframes=50, fps=4)
anim_save("animHomI4.gif", p)


# subjects
set.seed(123)
pat1 <- tibble(
  r0l = rnorm(nPop, log(r0b), sd0b), # log of patient specific rate
  r0 = exp(r0l) / exp(sd0b^2/2) # patient specific rate (rescaled to have mean r0)
)

set.seed(123)
sim1 <- mySim(pat1)

R360 = sim1 %>% summarise(mr = mean(r0 * (is == 0)) * (idur - 1)) %>% unlist()

sim1 <-
  sim1 %>%
  mutate(
    iweek = pmin((is + 6) %/% 7, 52))

pp1 <-
  ggplot(sim1, aes(x=r0)) + 
  # geom_freqpoly(fill = NA, size=2) +
  geom_histogram(aes(fill = (is>0)), bins=40, alpha=0.2) +
  geom_freqpoly(aes(colour=factor(iweek)), position="stack", bins=40) + 
  scale_x_log10() +
  xlab("Infection rate (1/Day)") +
  scale_fill_discrete(labels = c("No", "Yes")) +
  guides(colour = "none", fill = guide_legend("Infected")) +
  ggtitle(paste0("Heterogeneous\nR(1 year) = ", round(R360, 2)))


print(pp1)

p <- mygraph1(sim1)
animate(p, nframes=50, fps=4)
anim_save("animHetBase.gif", p)


set.seed(123)
sim1 <- mySim(pat1, int4)

R360 = sim1 %>% summarise(mr = mean(r0 * (is == 0)) * (idur - 1)) %>% unlist()

sim1 <-
  sim1 %>%
  mutate(
    iweek = pmin((is + 6) %/% 7, 52))

pp1 <-
  ggplot(sim1, aes(x=r0)) + 
  # geom_freqpoly(fill = NA, size=2) +
  geom_histogram(aes(fill = (is>0)), bins=40, alpha=0.2) +
  geom_freqpoly(aes(colour=factor(iweek)), position="stack", bins=40) + 
  scale_x_log10() +
  xlab("Infection rate (1/Day)") +
  scale_fill_discrete(labels = c("No", "Yes")) +
  guides(colour = "none", fill = guide_legend("Infected")) +
  ggtitle(paste0("Heterogeneous with follow up to day 180\nR(1 year) = ", round(R360, 2)))


print(pp1)

p <- mygraph1(sim1)
animate(p, nframes=50, fps=4)
anim_save("animHetI4.gif", p)


```



